---
title: "lab10"
format:
  html:
    embed-resources: true
editor: visual
---

## 1

```{r}
library(ggplot2)
library(plotly)


# load COVID state-level data from NYT
cv_states <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

# load state population data
state_pops <- read.csv("https://raw.githubusercontent.com/COVID19Tracking/associated-data/master/us_census_data/us_census_2018_population_estimates_states.csv")

# merge datasets
cv_states <- merge(cv_states, state_pops, by.x = "state", by.y = "state_name")

```

## 2

```{r}

dim(cv_states)

head(cv_states)

tail(cv_states)

str(cv_states)
```

## 3

```{r}
# format the date
cv_states$date <- as.Date(cv_states$date)

# format the state and state abbreviation (abb) variables
cv_states$state <- factor(cv_states$state)
cv_states$abb <- factor(cv_states$state.y)

# order the data first by state, second by date
cv_states <- cv_states[order(cv_states$state, cv_states$date),]

# Confirm the variables are now correctly formatted
str(cv_states)
head(cv_states)
tail(cv_states)

# Inspect the range values for each variable.
head(cv_states)
summary(cv_states)
min(cv_states$date)
max(cv_states$date)
```

## 4

```{r}
library(dplyr)
library(ggplot2)
library(plotly)
library(zoo)

state_list <- unique(cv_states$state)

# Add variables for new_cases and new_deaths:
for (i in 1:length(state_list)) {
  cv_subset <- subset(cv_states, state == state_list[i])
  cv_subset <- cv_subset[order(cv_subset$date), ]

  # add starting level for new cases and deaths
  cv_subset$new_cases <- cv_subset$cases[1]
  cv_subset$new_deaths <- cv_subset$deaths[1]

  for (j in 2:nrow(cv_subset)) {
    cv_subset$new_cases[j] <- cv_subset$cases[j] - cv_subset$cases[j - 1]
    cv_subset$new_deaths[j] <- cv_subset$deaths[j] - cv_subset$deaths[j - 1]
  }

  # include in main dataset
  cv_states$new_cases[cv_states$state == state_list[i]] <- cv_subset$new_cases
  cv_states$new_deaths[cv_states$state == state_list[i]] <- cv_subset$new_deaths
}

# Focus on recent dates
cv_states <- cv_states |> dplyr::filter(date >= "2021-06-01")

# Inspect implausible values in new_cases using plotly
weird_cases <- cv_states |>
  dplyr::filter(new_cases < 0 | new_cases > quantile(new_cases, 0.995, na.rm = TRUE))

p1 <- ggplot(
  weird_cases,
  aes(
    x = date,
    y = new_cases,
    text = paste(
      "State:", state,
      "<br>Date:", date,
      "<br>New cases:", new_cases
    )
  )
) +
  geom_point(size = 1, alpha = 0.7)

ggplotly(p1, tooltip = "text")

# Inspect implausible values in new_deaths using plotly
weird_deaths <- cv_states |>
  dplyr::filter(new_deaths < 0 | new_deaths > quantile(new_deaths, 0.995, na.rm = TRUE))

p2 <- ggplot(
  weird_deaths,
  aes(
    x = date,
    y = new_deaths,
    text = paste(
      "State:", state,
      "<br>Date:", date,
      "<br>New deaths:", new_deaths
    )
  )
) +
  geom_point(size = 1, alpha = 0.7)

ggplotly(p2, tooltip = "text")

# set negative new case or death counts to 0
cv_states$new_cases[cv_states$new_cases < 0] <- 0
cv_states$new_deaths[cv_states$new_deaths < 0] <- 0

# Re-calculate cases and deaths as cumulative sum of updated new_cases and new_deaths
for (i in 1:length(state_list)) {
  cv_subset <- subset(cv_states, state == state_list[i])
  cv_subset <- cv_subset[order(cv_subset$date), ]

  # add starting level for new cases and deaths
  cv_subset$cases[1] <- cv_subset$new_cases[1]
  cv_subset$deaths[1] <- cv_subset$new_deaths[1]

  for (j in 2:nrow(cv_subset)) {
    cv_subset$cases[j] <- cv_subset$new_cases[j] + cv_subset$cases[j - 1]
    cv_subset$deaths[j] <- cv_subset$new_deaths[j] + cv_subset$deaths[j - 1]
  }

  # include in main dataset
  cv_states$cases[cv_states$state == state_list[i]] <- cv_subset$cases
  cv_states$deaths[cv_states$state == state_list[i]] <- cv_subset$deaths
}

# Smooth new counts
cv_states$new_cases <- zoo::rollmean(cv_states$new_cases, k = 7, fill = NA, align = "right") |> round(0)
cv_states$new_deaths <- zoo::rollmean(cv_states$new_deaths, k = 7, fill = NA, align = "right") |> round(0)

# Inspect data again interactively
recent_deaths <- cv_states |>
  dplyr::filter(new_deaths > quantile(new_deaths, 0.995, na.rm = TRUE))

p2 <- ggplot(
  recent_deaths,
  aes(
    x = date,
    y = new_deaths,
    text = paste(
      "State:", state,
      "<br>Date:", date,
      "<br>Smoothed new deaths:", new_deaths
    )
  )
) +
  geom_point(size = 1, alpha = 0.7)

ggplotly(p2, tooltip = "text")
```

**In both variables, there are many negative values, which are not meaningful for daily counts. These negative values likely result from retrospective data corrections or reporting adjustments.**

**Additionally, we observed unusually large spikes in both new cases and new deaths. These spikes may reflect delayed reporting or backlog releases.**

**After correcting the negative values by setting them to zero, we recalculated the cumulative counts to ensure consistency.**

## 5

```{r, warning=FALSE}
# add population normalized (by 100,000) counts for each variable
cv_states$per100k <- as.numeric(format(round(cv_states$cases/(cv_states$population/100000), 1), nsmall = 1))
cv_states$newper100k <- as.numeric(format(round(cv_states$new_cases/(cv_states$population/100000), 1), nsmall = 1))
cv_states$deathsper100k <- as.numeric(format(round(cv_states$deaths/(cv_states$population/100000), 1), nsmall = 1))
cv_states$newdeathsper100k <- as.numeric(format(round(cv_states$new_deaths/(cv_states$population/100000), 1), nsmall = 1))

# add a naive_CFR variable = deaths / cases
cv_states <- cv_states |> mutate(naive_CFR = (deaths / cases))

# create a `cv_states_today` variable
cv_states_today <- subset(cv_states, date == max(cv_states$date))
```

## 6

```{r,warning=FALSE}
# pop_density vs. cases
cv_states_today |> 
  plot_ly(x = ~pop_density, y = ~cases, 
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70), marker = list(sizemode='diameter', opacity=0.5))

# filter out one outlier state
cv_states_today_filter <- cv_states_today |> filter(state!="District of Columbia")

# pop_density vs. cases after filtering
cv_states_today_filter |> 
  plot_ly(x = ~pop_density, y = ~cases, 
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70), marker = list(sizemode='diameter', opacity=0.5))

# pop_density vs. deathsper100k
cv_states_today_filter |> 
  plot_ly(x = ~pop_density, y = ~deathsper100k,
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70), marker = list(sizemode='diameter', opacity=0.5))

# Adding hoverinfo
cv_states_today_filter |> 
  plot_ly(x = ~pop_density, y = ~deathsper100k,
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70), marker = list(sizemode='diameter', opacity=0.5),
          hoverinfo = 'text',
          text = ~paste(paste(state, ":", sep=""),
                        paste(" Cases per 100k: ", per100k, sep=""),
                        paste(" Deaths per 100k: ", deathsper100k, sep=""),
                        sep = "<br>")) |>
  layout(title = "Population-normalized COVID-19 deaths (per 100k) vs. population density for US states",
         yaxis = list(title = "Deaths per 100k"),
         xaxis = list(title = "Population Density"),
         hovermode = "compare")
```

## 7

```{r,warning=FALSE}
p <- ggplot(cv_states_today_filter, 
            aes(x = pop_density, y = deathsper100k, size = population)) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE, color = "blue")

ggplotly(p)
```

## 8

```{r}
MULT_FACTOR <- 50

cv_states |> 
  filter(state == "Florida") |> 
  arrange(date) |> 
  plot_ly(x = ~date, y = ~new_cases / MULT_FACTOR, type = "scatter", mode = "lines", name = "Cases") |> 
  add_trace(x = ~date, y = ~new_deaths, type = "scatter", mode = "lines", name = "Deaths")
```

## 9

```{r}
# Map state, date, and new_cases to a matrix
library(tidyr)
cv_states_mat <- cv_states |> select(state, date, new_cases) |> dplyr::filter(date > as.Date("2021-06-15"))
cv_states_mat2 <- as.data.frame(pivot_wider(cv_states_mat, names_from = state, values_from = new_cases))
rownames(cv_states_mat2) <- cv_states_mat2$date
cv_states_mat2$date <- NULL
cv_states_mat2 <- as.matrix(cv_states_mat2)

# Create a heatmap using plot_ly()
plot_ly(x = colnames(cv_states_mat2), y = rownames(cv_states_mat2),
        z = ~cv_states_mat2,
        type = "heatmap",
        showscale = TRUE)

# Repeat with newper100k
cv_states_mat <- cv_states |> select(state, date, newper100k) |> dplyr::filter(date > as.Date("2021-06-15"))
cv_states_mat2 <- as.data.frame(pivot_wider(cv_states_mat, names_from = state, values_from = newper100k))
rownames(cv_states_mat2) <- cv_states_mat2$date
cv_states_mat2$date <- NULL
cv_states_mat2 <- as.matrix(cv_states_mat2)

plot_ly(x = colnames(cv_states_mat2), y = rownames(cv_states_mat2),
        z = ~cv_states_mat2,
        type = "heatmap",
        showscale = TRUE)

# Create a second heatmap after filtering to only include dates every other week
filter_dates <- seq(as.Date("2021-06-15"), as.Date("2021-11-01"), by = "2 weeks")

cv_states_mat <- cv_states |> select(state, date, newper100k) |> filter(date %in% filter_dates)
cv_states_mat2 <- as.data.frame(pivot_wider(cv_states_mat, names_from = state, values_from = newper100k))
rownames(cv_states_mat2) <- cv_states_mat2$date
cv_states_mat2$date <- NULL
cv_states_mat2 <- as.matrix(cv_states_mat2)

# Create a heatmap using plot_ly()
plot_ly(x = colnames(cv_states_mat2), y = rownames(cv_states_mat2),
        z = ~cv_states_mat2,
        type = "heatmap",
        showscale = TRUE)
```
