---
title: "Lab 4 - Data Visualization"
link-citations: true
toc: false
embed-resources: true
---

```{r, echo = FALSE}
# Set to TRUE if you want the blocks to be evaluated 
EVAL <- FALSE
ECHO <- EVAL
```

```{r setup, message=FALSE, echo = ECHO, warning=FALSE, eval = EVAL}
library(tidyverse)
```

# Learning Goals

-   Read in and prepare the COVID-19 dataset
-   Create several graphs with different `geoms()` in `ggplot2`
-   Create a facet graph
-   Customize your plots
-   Create a detailed map

# Lab Description

In this lab, we will work with the COVID-19 data from the CDC. The data has already been preprocessed.

**The objective of the lab is to examine the association between weekly average COVID-19 death and vaccination rates in ten regions of the US.**

Note the following variable definitions:

-   `mmwr_year`: year (according to the [MMWR Calendar](https://stacks.cdc.gov/view/cdc/22305))
-   `mmwr_week`: week (according to the [MMWR Calendar](https://stacks.cdc.gov/view/cdc/22305))
-   `pop`: state population
-   `cases`: weekly COVID cases
-   `hosp`: weekly COVID hospitilizations
-   `deaths`: weekly COVID deaths
-   `series`: cumulative vaccinated population
-   `booster`: cumulative boosted population

# Steps

### 1. Read in the data

First download and then read in with `read.csv()`

```{r, echo = TRUE, message=FALSE, eval = EVAL}
if (!file.exists("covid_processed.csv"))
  download.file(
    url = "https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/covid/covid_processed.csv",
    destfile = "covid_processed.csv",
    method   = "libcurl",
    timeout  = 60
    )
```

### 2. Prepare the data

-   Since states have different population sizes, make sure that you are working with population-normalized rates (e.g. units cases per 100,000 population). Transform the following variables accordingly: `cases`, `hosp`, `booster`, `series`, `deaths`. 
-   Use the `ymd` function from the `lubridate` to convert the date variable to a Date object. This is easier to work with for arithmetic and plotting.
-   Compute the mean by state of the normalized variables above. We recommend creating a new data.frame with the state-averaged data. Keep any other relevant variables from the original data (e.g. population, region etc).
-   Create a logical variable for low vs high population.
-   Make sure that region is a factor

```{r, eval = EVAL, echo = ECHO}
### Your code here
```

### 3. Use `geom_violin` to examine the case rates and death rates by region

You saw how to use `geom_boxplot` in class. Try using `geom_violin` instead (take a look at the help). (hint: you will need to set the `x` aesthetic to 1)

-   Use facets
-   Fill color by region
-   Describe what you observe in the graph

```{r, eval = EVAL, echo = ECHO}
### Your code here
```

### 4. Use `geom_point` with `stat_smooth` to examine the association between time and case rates by region

-   Filter out the data to exclude the first two weeks in the dataset
-   Color points by region
-   Fit a function with `stat_smooth` by region. Does `method=lm` or the default method work better?
-   For the default `stat_smooth`, play around with the `span` parameter to get a smooth that fits the data better without being too noisy.
-   Describe what you observe in the graph

```{r, eval = EVAL, echo = ECHO}
### Your code here
```

### 5. Use `geom_bar` to create barplots of the states by population category colored by region

-   Bars by population category using `position="dodge"`
-   Change colors from the default. Color by region using `scale_fill_brewer` see [this](http://rstudio-pubs-static.s3.amazonaws.com/5312_98fc1aba2d5740dd849a5ab797cc2c8d.html)
-   Create nice labels on the axes and add a title
-   Describe what you observe in the graph

```{r warning=FALSE, message=FALSE, eval = EVAL, echo = ECHO}
### Your code here
```

### 6. Use `stat_summary` to examine mean vaccination rate and death rate by region with standard deviation error bars

-   Use `fun.data="mean_sdl"` in `stat_summary`. Look up what `mean_sdl` does.
-   Add another layer of `stats_summary` but change the geom to `"errorbar"` (see the help).
-   Describe the graph and what you observe

```{r, eval = EVAL, echo = ECHO}
### Your code here
```

-   Vaccination rates are ...
-   Death rates are...

### 7. Make a map showing the spatial trend in COVID deaths in the US

-   Use `geom_polygon` and `map_data`. Modify the given code.
-   Merge the `map_data` data.frame with your COVID data.frame to add in the COVID data to the map data. Hint: use `left_join`.
-   Use a color palette with custom colors
-   Make sure that your legend is labelled correctly.
-   Add a * or other label for the top 10 states in highest death rate.
-   Make sure that the aspect ratio is appropriate
-   Describe your observations

```{r, eval = EVAL, echo = ECHO}
us_map <- map_data("state")
ggplot(data = us_map, aes(x = long, y = lat, fill = region)) + geom_polygon(color = "white") 

### Your code here
```

-   Describe the trend in COVID death rates across the US

### 8. Use a ggplot extension

-   Pick an extension (except `cowplot`) from [here](https://exts.ggplot2.tidyverse.org/gallery/) and make a plot of your choice using the COVID data
-   You might want to try examples that come with the extension first (e.g. `ggtech`, `gganimate`, `ggforce`)

### 9. Submission (due Tuesday, Feb 10 at 8:30am)

Submit through the course [Github issues page](https://github.com/dmcable/BIOSTAT620W26/issues/3).
