---
title: "lab8"
format:
  html:
    embed-resources: true
editor: visual
---

## 1.1

```{r}
library(microbenchmark)
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n) {
    x <- rbind(x, rpois(k, lambda))
  }
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
  matrix(rpois(n * k, lambda), nrow = n, ncol = k)
}

set.seed(555)
x1 <- fun1()
x2 <- fun1alt()

dim(x1)
dim(x2)

mean(x1)
mean(x2)

var(as.vector(x1))
var(as.vector(x2))


microbenchmark(
  fun1 = fun1(n = 1000, k = 20, lambda = 4),
  fun1alt = fun1alt(n = 1000, k = 20, lambda = 4),
  times = 100
)
```

## 1.2

```{r}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max)
}

fun2alt <- function(x) {
  out <- x[1, ]
  for (i in 2:nrow(x)) {
    out <- pmax(out, x[i, ])
  }
  out
}

y1 <- fun2(x)
y2 <- fun2alt(x)

identical(y1, y2)
all.equal(y1, y2)

microbenchmark(
  fun2 = fun2(x),
  fun2alt = fun2alt(x),
  times = 100
)
```

## 2.1

```{r}
library(parallel)

my_boot <- function(dat, stat, R, ncpus = 1L) {

  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n * R, TRUE), nrow = n, ncol = R)

  cl <- makeCluster(ncpus)
  clusterExport(cl, varlist = c("dat", "idx", "stat"), envir = environment())

  ans <- parLapply(cl, seq_len(R), function(i) {
    stat(dat[idx[, i], , drop = FALSE])
  })

  stopCluster(cl)

  # Convert list to matrix
  ans <- do.call(rbind, ans)

  return(ans)
}


```

## 2.2

```{r}
# Bootstrap of an OLS
my_stat <- function(d) coef(lm(y ~ x, data=d))

# DATA SIM
set.seed(1)
n <- 500; R <- 1e4

x <- cbind(rnorm(n)); y <- x*5 + rnorm(n)

# Checking if we get something similar as lm
ans0 <- confint(lm(y~x))
ans1 <- my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 2L)

# You should get something like this
t(apply(ans1, 2, quantile, c(.025,.975)))
##                   2.5%      97.5%
## (Intercept) -0.1372435 0.05074397
## x            4.8680977 5.04539763
ans0
##                  2.5 %     97.5 %
## (Intercept) -0.1379033 0.04797344
## x            4.8650100 5.04883353
```

## 2.3

```{r,warning=FALSE}
Sys.setlocale("LC_ALL", "English")
system.time(
  my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 1L)
)


system.time(
  my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 2L)
)


system.time(
  my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 4L)
)
```

## 3.1

A problem is suitable for parallelization when the computation can be divided into many independent tasks that do not depend on each other. Each task can then be processed on a different CPU core at the same time.

Vectorization and parallelization are both ways to make code faster. Vectorization speeds up computation by using optimized operations on entire vectors or matrices instead of loops. Parallelization speeds up computation by distributing tasks across multiple CPU cores. Usually best to first use vectorization if possible, since it is simpler and often very efficient, and then use parallelization when the task still takes a long time and can be split into independent parts.
